Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-10-27 paper/main.tex
work on the new speach

Static analysis of quantitative properties of imperative
programs is an active research area and recent years
have seen many innovations.  Notable tools that
have been developed include SPEED~\cite{GulwaniMC09},
KoAT~\cite{BrockschmidtEFFG14}, PUBS~\cite{AlbertAGPZ12},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.

These tools can derive impressive results on realistic
system software.  However, there are still theoritical
shortcomings that remain open problems to this day:

\begin{itemize}
\item The computed bounds are often
  \emph{non-compositional} local bounds (for a single
  (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item When an automatic tool fails on a loop of the
  program, no user interaction is possible to derive
  a bound on the full program.
\itemskipIn
\item Existing tools work on a highly abstracted
  version of the program, thus, are unable to provide
  a simple syntactic proof of the correctness of their
  bound.
\itemskipIn
\item Maybe we can mention resource freeing                    XXX
\end{itemize}

This paper presents a new framework to derive worst-case
resource bounds for C programs.  Our choice of C is
mainly motivated by its strong presence in system and
embedded software.  These applications often have to meet
strict time and memory constraints.  To ensure that these
requirements are met, profiling is not enough because non
exhaustive.  In contrast, static analysis tools provide
stronger and sound guarantees.

Our compositional framework is built of three parts: 1-
a clear cost model for C programs based on instrumented
semantics; 2- a quantitative Hoare logic allowing
compositional reasoning on the resource consumption of
C programs at a syntactic level; 3- an automated tool
that derives bounds in the previous logic.

Compared to existing work, our framework is a jump in
compositionality and correctness: the analysis is no
longer centered on single loops but on complete programs,
a mix of automation and interaction can be used to derive
global bounds, and bounds are computed and proved within
a clear syntactic deductive system.  Compositionality is
due first to the logic that soundly allows mixes of hand
derived bounds and automatically generated bounds.  It is
also due to the automated tool that keeps track of loop
bounds and size changes of variables in a unified system.

The cost model and the quantitative Hoare logic
presented are based on a recent work~\cite{veristack14}
that formally verifies bounds on the stack usage of
C programs.  We generalize this work by allowing to
reason on any resource.  The automation, which is the
keystone of this paper, is based on amortized analysis
techniques~\cite{jan-something}.  It makes the analysis
of large programs tractable by automatically generating
resource bounds in the previous logic on a large class
of programs.

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers and
differences of variables.  It is also the first resource
analysis tool for C that deals naturally with recursive
functions, resources that can be freed (e.g. memory),
and sequenced loops.  Automated tools for functional
programs already existed~\cite{Jost03,HoffmannAH12}
but it has been a long-time open problem to apply it to
C-like programming languages with control flow depending
on integer arithmetic.

Compared to more classical approaches based on ranking
functions, tools that are based on amortized analysis
only need one mechanism to provide features that are
usually treated non uniformly:

\begin{itemize}
\item interaction between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\item and \emph{amortized bounds} as found
  in the Knuth-Morris-Pratt algorithm for string
  search.
\end{itemize}

This paper presents a compositional framework for
deriving certified resource bounds on C programs. We
emphasize the following contributions: 1- an automatic
analysis that generates linear compositional and
certified resource bounds; 2- an infrastructure verified
in Coq that lets a programmer derive complex free-form
resource bounds on complete C programs.

>>>>

% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW

Available techniques split in two categories.  In the
first one, tools derive impressive bounds for
imperative programs, but are not fully compositional.
Notable recent work like
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14} falls in this category.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are expressed
as arithmetic expressions in terms of variables as they
are just before the loop.  This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
XXX MENTION GULWANI PAPER [1] XXX
Second, while all imperative programming languages
provide a function or procedure abstraction, all
available tools are not able to abstract their resource
behavior and have to inline the procedure body to perform
any kind of analysis.  Third, when such a tool fails on
given program, the only way out to get a resource bound
is to rewrite the program.  There is no possibility
of sound interaction with the tool, and manual proofs
cannot be composed with automatically inferred bounds.
One of the biggest challenge in the way of interaction
with these automated tools is that they work on a highly
abstract version of the program that has a lot more
behaviors than the original program.

In the second category, tools also derive very tight
bounds on functional programs~\cite{Jost03,HoffmannAH12}.
These tools are highly compositional, scale for large
programs, and work directly on the syntax.  Despite all
these theoretical and practical advantages, it has been
a long-time open problem how to make them work in the
presence of negative integers, mutation and non-linear
control flow.  For example, the first challenge of
imperative programs is the absence of useful type system
to hook the analysis on.  Another issue is that these
tools associate a \emph{potential} directly to program
variables, doing this in imperative programs would
fail for loops as simple as \lstinline{for(i=x;i<y;i++)}
where $y - i$ decreases, but not $|i|$.
Mutation is yet another challenge, while functional
programs relate multiple fixed inputs (the typing
context) with one output (the expression typed),
imperative programs can change the value of variables
in scope, so the potential of these variables has to be
updated after mutations or unsound bounds can be derived.

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: it solves the compositionality issues of
imperative world by adapting amortized-analysis based
techniques from the functional world.

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers
and differences of variables.  It is also the first
resource analysis tool for C that deals naturally with
recursive functions, resources that can be freed (e.g.
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning and can handle using
only one simple mechanism:

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}

The main innovations that make amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and track multivariate quantities instead
of program variables directly.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.

% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW
